
\section{Conservation of Angular Momentum}

\makelabheader %(Space for student name, etc., defined in master.tex or labmanual_formatting_commands.tex)

\textbf{Objectives} 

To test the Law of Conservation of Angular Momentum and to explore the applicability
of angular momentum conservation among objects that experience no external torques. 

\textbf{Apparatus}

\begin{itemize}
\item A Rotating Disk System 
\item Masses of 500 g, 1 kg 
\item A meter stick and a ruler 
\item A small water bubble level
\item A video analysis system (\textit{VideoPoint} or \textit{Tracker})
\end{itemize}
\textbf{Overview }

As a consequence of Newton's laws, angular momentum like linear momentum is
believed to be conserved in isolated systems. This means that, no matter how
many internal interactions occur, the total angular momentum of any system should remain constant if there are no external torques. When one of the objects gains some angular momentum another part of the system must lose the same amount. If angular momentum isn't conserved, then we believe that there is some outside torque acting on the system. By expanding the boundary of the system to include the source of that torque we can always preserve the Law of Angular Momentum Conservation. 

In this unit you will test the notion of the conservation of angular momentum.
As in the test of the conservation of linear momentum, we will investigate what
happens when two bodies undergo a ``rotational'' collision.
You will drop a large weight onto a rotating disk and determine the moment of
inertia, the angular speed, and finally, the angular momentum of the rotator-disk-weight
system before and after this perfectly inelastic collision.

\textbf{Activity 1: The Moment of Inertia Before and After the Collision}

(a) Calculate the theoretical value of the moment of inertia of the disk
using basic measurements of its radius and mass. Be sure to state units and
show the expression you used!
\vspace{5mm}

\( r_{d} =\)  \hfill{}\( M_{d}= \) \hfill{}
\vspace{5mm}

\( I_{d}= \)
\vspace{5mm}

(b) After dropping the weight on the rotating disk, the system will have a new
moment of inertia. Derive a formula for the moment of inertia of a cylindrical-shaped weight of mass \( m_{w} \) and radius \( r_{w} \) revolving about the origin at a distance \( r_{r} \). (You will have to use the parallel axis theorem to do this.)
\vspace{5mm}

\( I_{w} =\)  
\vspace{10mm}

(c) Record the mass of one of the weights and use a vernier caliper to measure its diameter. Obtain its radius from the diameter.
\vspace{5mm}

\( m_{w} =\)  \hfill{}\(r_{w} =\) \hfill{}
\vspace{5mm}

(d) Come up with a formula for the moment of inertia $I$ of the whole system
before and after the collision and calculate the moment of inertia before the
collision only. (The moment of inertia after the collision will be determined AFTER you do the experiment, because you will not know \( r_{r} \) until after you do the experiment.) Don't forget to include the units in \( I_{before} \).
\vspace{5mm}

\( I_{\mbox{\small before}}= \) 
\vspace{5mm}

\( I_{\mbox{\small after}} =\)  
\vspace{5mm}

\textbf{Activity 2: Making a Movie of the Collision} 

(a) Place the video camera about 1 m above the rotator, align the camera with
the center of the rotator using the pendulum, and center the rotator in the
field of view of the camera by viewing it with the \textit{VideoPoint
Capture} software.
Place a ruler of known length in the field of view of the camera and parallel
to one side of the frame. Check that the rotator is flat with the small water-bubble level.

(b) Give the rotator a push and begin recording its motion with the video camera. See \textbf{Appendix \ref{videopoint}: Video Analysis Using VideoPoint} or \textbf{Appendix \ref{tracker}: Video Analysis Using Tracker} for details on making the movie. While the rotator is moving hold the weight you recorded above near the rim of the disk and close to, but not quite
touching, the surface of the moving disk. After at least one revolution
of the disk drop the mass onto the disk and record the motion of
the disk for at least one revolution afterward. 

(c) \textit{Before removing the weight from the disk}, determine the distance of the center of the weight you dropped from the center of the rotator \( r_{r} \). To do this, measure the distance from the center
of the rotator to the edge of the weight \( r_{edge} \) and use the result from Activity 1 part (c) for the radius of the weight \( r_{w} \). Calculate the distance from the origin to the center of the weight \( r_{r} \) (it is just the sum of \( r_{edge} \) and \( r_{w} \)). Use these results and those from Activity 1 part (d) to calculate the final moment of inertia.
\vspace{5mm}

\( r_{\mbox{\small edge}}  =\) \hfill{}\( r_{w} \) = \hfill{}\( r_{r} \) =\hfill{}
\vspace{5mm}

\( I_{\mbox{\small after}} =\)  
\vspace{5mm}

\textbf{Activity 3: Measurement of Angular Velocity}

Determine the angular speed before and after the collision.

\begin{enumerate}
\item Find the last frame before you dropped the weight on the rotator and click on
the position of the white marker on the metal disk. Now go backward through the film until the rotator has gone through one full rotation. Estimate how many frames there are in one rotation. (Step through the frames and count.) This will be called \( N_{\rm before}\). You also need to know the time between frames \( \Delta  t_{\rm frame}\), which is just the reciprocal of the capture rate in frames per second.
\[
N_{\mbox{\small before}}=\qquad \qquad \qquad \Delta t_{\rm frame}=\]
Calculate the time for one rotation (the period $T$ of rotation) before the collision and the angular speed in radians per second.
\[
T_{\mbox{\small before}}=\qquad \qquad \qquad \omega _{\mbox{\small before}}=\]
%\newpage
\item We now follow a similar procedure to determine the angular speed after the collision.
%Under the \textbf{Edit} Menu highlight \textbf{Clear All...} to get rid of your%
%previous results. 
Find the first frame after you dropped the weight on the rotator
and click on the position of the white marker. Now go forward through the film and estimate the number of frames in one full rotation.
\[
N_{\mbox{\small after}}=\qquad \qquad \qquad \Delta t_{\rm frame}=\]
Calculate the time for one rotation after the collision and the angular speed in radians per second.
\[
T_{\mbox{\small after}}=\qquad \qquad \qquad \omega _{\mbox{\small after}}=\]

\end{enumerate}
\vspace{20mm}

\textbf{Activity 4: Calculation of Angular Momentum}

(a) Calculate the angular momentum before and after the collision, including UNITS. Calculate the difference between the two results and record it below. 
Go around to the other lab groups and get their results for the difference between the angular momenta before and after the collision.
Make a histogram of the results you collect and calculate the average and standard deviation.
For information on making histograms, see \textbf{Appendix \ref{excel}}. For information on calculating the average and
standard deviation, see \textbf{Appendix \ref{treatment}}. Record the average and standard deviation here.
Attach the histogram to this unit.
\vspace{5mm}

\( L_{\mbox{\small before}}= \)  
\vspace{5mm}

\( L_{\mbox{\small after}}= \)
\vspace{5mm}

\( \Delta L= L_{\mbox{\small after}} - L_{\mbox{\small before}} = \)  
\vspace{15mm}

(b) What is your expectation for the difference between the initial and final angular momentum?
Do the data from the class support this expectation? 
Use the average and standard deviation for the class to quantitatively answer this question.
\vspace{20mm}

(c) What does the histogram of the class data tell you? Be quantitative in your answer.
\vspace{20mm}

(d) Within the limits of experimental uncertainty, is angular momentum 
conserved?  Be quantitative.
\vspace{20mm}

(e) Would the procedure you followed above change if the weight was moving horizontally at a constant velocity when you dropped it? 
If it changed, what would be different?
\vspace{20mm}

