
\section{Newton's Second Law for Rotation\footnote{
1990-93 Dept. of Physics and Astronomy, Dickinson College. Supported by FIPSE
(U.S. Dept. of Ed.) and NSF. Portions of this material may have been modified
locally and may not have been classroom tested at Dickinson College.
}}

\makelabheader %(Space for student name, etc., defined in master.tex or labmanual_formatting_commands.tex)

\textbf{Objectives} 

To understand torque and its relation to angular acceleration and moment of
inertia on the basis of both observations and theory. 

\textbf{Apparatus}

\begin{itemize}
\item A Rotating Disk System 
\item A hanging mass of 200 g (for applying torque) 
\item String and pulley
\item A movie scaling ruler and a simple cm ruler
\item Vernier caliper
\item Small water bubble level
\item A video analysis system (\textit{Tracker}).
\end{itemize}
\textbf{Overview} 

We have used the definition of moment of inertia, $I$, to determine a theoretical equation for the moment of inertia of a uniform disk. This equation was given by
\[
I=\frac{1}{2}MR^{2}.\]


Does this equation adequately describe the moment of inertia of a rotating
disk system? If so, then we should find that, if we apply a known torque, \( \tau  \), to the disk system, its resulting angular acceleration, \( \alpha  \), is actually related to the system's moment of inertia, $I$, by the equation
\[
\tau =I\alpha \quad \mbox{or}\quad \alpha =\tau /I\]


The purpose of this experiment is to determine if, within the limits of experimental
uncertainty, the measured angular acceleration of a rotating disk system is
the same as its theoretical value. The theoretical value of angular acceleration can be calculated using theoretically determined values for the torque on the
system and its moment of inertia.

\textbf{Theoretical Calculations} 

You'll need to take some basic measurements on the rotating disk system
to determine theoretical values for $I$ and \( \tau  \). Values of moment of
inertia calculated from the dimensions of a rotating object are theoretical
because they purport to describe the resistance of an object to rotation. An
experimental value is obtained by applying a known torque to the object and
measuring the resultant angular acceleration.

\newpage

\textbf{Activity 1: Theoretical Calculations }

(a) Calculate the theoretical value of the moment of inertia of the metal disk
using basic measurements of its radius and mass. Ignore the small hole in the
middle in your calculation (i.e. assume the disk is uniform). Be sure to state 
units!
\vspace{5mm}

\( R_{d} =\) \hfill{}\( M_{d}= \) \hfill{}
\vspace{5mm}

\( I = \) 
\answerspace{5mm}

(b) In preparation for calculating the torque on your system, summarize the
measurements for the falling mass, $m$, and the radius of the spool that has the string wrapped around it in the space below. The diameter of the spool can be measured using the vernier caliper. Don't forget the units!
\vspace{5mm}

$m = $\hfill{}\(r_{s}= \)\hfill{} 
\answerspace{5mm}

(c) Calculate the theoretical value for the torque on the rotating system as
a function of the magnitude of the hanging mass and the radius, \( r_{s} \),
of the spool, assuming the tension in the string is equal to the weight of the 
falling mass (this introduces an error of less than 1 percent if $m$ = 200 g). 
Remember that torque equals force (tension in the string) times 
moment arm (radius of spool). Be sure to include units.
\vspace{5mm}

\( \tau _{th}= \)
\answerspace{5mm}

(d) Based on the values of torque and moment of inertia that you just 
calculated, what is the theoretical value of the angular acceleration of the 
disk? (Refer to equations in \textbf{Overview}). Be sure to include units.
\vspace{5mm}

\( \alpha _{th}= \)
\answerspace{5mm}

\textbf{Activity 2: Experimental Measurement of Angular Acceleration} 

(a) Place the video camera about 1 m above the rotator, and align the camera with the 
center of the rotator. Open \textbf{Camera} and turn on the camera as explained in \textbf{Appendix \ref{tracker}: Video analysis Using Tracker}. Center the rotator in the 
field of view of the camera. Use the small level to ensure that the surface of the rotator 
is level. Place a ruler of known length in the field of view of the camera and parallel to 
one side of the frame.

(b) Place the rotator so the string will pass smoothly over the pulley and put
200 g of mass on the end of the string. Release the rotator and use the
video camera to record the motion of the disk for at least two full turns. See  
\textbf{Appendix \ref{tracker}: Video analysis Using Tracker} for details on 
making the movie.

(c) Determine the angular displacement of the rotator as a function of time. \textbf{Important:} Be careful to place the origin of your coordinate system on the axle of the rotator BEFORE recording data, so the angular displacement you measure will be the desired one. Using \textit{Tracker}, track the outer edge of the white marker on the rotating disk for at least two full rotations. The resulting file should contain three columns with the values of time, $x$-position, and $y$-position for two complete rotations of the disk.

\pagebreak[2]
(d) What is the expression for the angular displacement of the disk (relative to the $x$ axis) in terms of the $x$- and $y$-positions of the marker that you recorded above? (Note that these positions will be relative to an origin that you placed on the axle of the rotator in part (c).)
\vspace{5mm}

\( \theta  \)= 
\bigskip

\newpage
(e) We want to graph the angular displacement of the disk as a function of time.
To do this:

\begin{enumerate}
\item Click on the word {\it Table} immediately above your data in Tracker and select \( \theta \) in the Visible Table Columns window that pops up.  A new column of data labeled \( \theta \) should appear.

\item Next {\bf right} click on the \( \theta \) column label in the data table and select Number and Units.  At the bottom of the window that pops up select radians. 

\item  Use this new data of the angular displacement to plot the angular displacement as a function of time. 

\end{enumerate}
(f) We now want to extract the angular acceleration from the data.

\begin{enumerate}
\item To describe the time dependence of the angular displacement what type of 
function should we use to fit the data? (Refer to Activity 6 part (d) of the lab Introduction to Rotation.) Which coefficient of the polynomial is related to 
the angular acceleration?
\answerspace{20mm}


\item Add a trendline to the data with a polynomial fit of order 2 and display the resulting equation for the time dependence on the graph (see Appendix B for help with doing this in Tracker or export your data to Excel). Write the equation for the angular position as a function of time in the space below. Be sure to include the proper units with the coefficients. Determine the experimental value for the angular acceleration from the fit and record it below. 
\vspace{20mm}

\end{enumerate}


\textbf{Activity 3: Comparing Theory with Experiment }

(a) Summarize the theoretical and experimental values of angular acceleration. 
Be sure to include the proper units.
\vspace{10mm}

\( \alpha _{\rm th}= \)
\vspace{10mm}

\( \alpha _{\rm exp}= \) 
\vspace{15mm}

(b) Do theory and experiment agree within the limits of experimental 
uncertainty? Calculate the percent difference.

